Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2019-02-05 histogram.tex
Upload a new refined version of the manuscript prepared for this journal

\author{Rodrigo~Ramele,~\IEEEmembership{Member,~IEEE,}
        Ana~Julia~Villar,
        and~Juan~Miguel~Santos% <-this % stops a space
\thanks{R. Ramele, A.J.Villar and J.M.Santos are with the Department
of  Computer Engineering, Instituto Tecn{o'}gico de Buenos Aires (ITBA), Ciudad de Buenos Aires,  Argentina e-mail: rramele@itba.edu.ar}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The analysis of Electroencephalographic (EEG) signals is of ulterior importance to aid in the diagnosis of mental disease and to increase our understanding of the brain.  Traditionally, clinical EEG has been analyzed in terms of temporal waveforms, looking at rhythms in spontaneous activity, subjectively identifying troughs and peaks in Event-Related Potentials (ERP), or by studying graphoelements in pathological sleep stages.
Additionally, the discipline of Brain Computer Interfaces requires new methods to decode patterns from non-invasive EEG signals.  This field is developing alternative communication pathways to transmit volitional information from the Central Nervous System. The technology could potentially enhance the quality of life of patients affected by neurodegenerative disorders and other mental illness. This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from images of signal plots.  These features are constructed based on the calculation of histograms of oriented gradients from pixels around the signal plot. It aims to provide a new objective framework to analyze, characterize and classify EEG signal waveforms. The feasibility of the method is outlined by detecting the P300, an ERP elicited by the oddball paradigm of rare events, and implementing an offline P300-based BCI Speller.  The validity of the proposal is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset of healthy subjects.
\end{abstract}

@@

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related to the image scale. In addition, $ x^l(n,c) $ is the point-to-point averaged multichannel EEG signal for the sample point $n$ and for channel $c$. Lastly, $$\bar{x}^l(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x^l(n,c)$$ and $$ \hat{\sigma}^l(c) = (\frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}}(x^l(n,c)-\bar{x}^l(c))^2 )^{\frac{1}{2}}$$ are the mean and estimated standard deviation of $x^l(n,c), 1 \leq n \leq n_{max}$, for each channel $c$.

@@

\begin{equation}
I^{(l,c)}(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \   n \quad  \text{and} \quad  z_2 = \tilde{x}^l(n,c) + z^l(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

@@

\begin{align*}
z^l(c) = \left \lfloor{ \min_{n} \tilde{x}^l(n,c)  }\right \rfloor
\label{eq:zerolevel}
\end{align*}

@@

In Table~\ref{tab:resultsown} the results obtained for 8 healthy subjects are shown.  It can be observed that the performance is above chance level. It was verified that HIST method has an improved performance at letter identification than SVM that process the signals on a channel by channel strategy (Wilcoxon signed-rank test, $p =  0.004$ for both datasets).

@@

Tables~\ref{tab:resultsalsswlda} and~\ref{tab:resultsownswlda} are presented in order to compare the performance of the HIST method versus a multichannel version of the Stepwise Linear Discriminant Analysis (SWLDA) and SVM classification algorithms for both datasets.  The feature was formed by concatenating all the channels~\cite{Krusienski2006}.  SWLDA is the methodology proposed by the ALS dataset's publisher. Since authors \cite{Riccio2013} did not report the Character Recognition Rate obtained for this dataset, we replicate their procedure and include the performance obtained with the SWLDA algorithm at letter identification.  It was verified for the dataset of ALS patients that it has similar performance  against other methods like SWLDA or SVM, which use a multichannel feature (Quade test with $p=0.55$) whereas for the dataset of healthy subjects significant differences were found (Quade test with $p=0.02$) where only the HIST method achieved a different performance than SVM (with multiple comparisons, significant difference of level $0.05$).

@@

\begin{figure*}[!t]
\centering
\subfloat[Case I]{\includegraphics[width=15cm]{subject.png}\label{subject8}}
\caption{Ten sample P300 template patches for subjects 8 (A) and 3 (B) of the ALS Dataset.  Downward deflection is positive polarity. }
\label{fig:p300templates}
\end{figure*}

Additionally, the stability of the P300 component waveform has been extensively studied in patients with ALS \cite{SellersandEmanuelDonchin2006,TomohiroMadarame2008,Nijboer2009,Mak2012,McCane2015} where it was found that these patients have a stable P300 component, which were also sustained across different sessions.  In line with these results we do not find evidence of a difference in terms of the performance obtained by analyzing the waveforms (HIST) for the group of patients with ALS and the healthy group of volunteers (Mann-Whitney U Test, $p=0.46$). Particularly, the best performance is obtained for a subject from the ALS dataset for which, based on visual observation, the shape of they P300 component is consistently identified.

@@

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}{Michael Shell}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{John Doe}
Biography text here.
\end{IEEEbiographynophoto}

@@

\begin{IEEEbiographynophoto}{Jane Doe}
Biography text here.
\end{IEEEbiographynophoto}

>>>>

\author{Rodrigo~Ramele,~\IEEEmembership{Member,~IEEE,}
        Ana~Julia~Villar,
        and~Juan~Miguel~Santos% <-this % stops a space
\thanks{R. Ramele, A.J.Villar and J.M.Santos are with the Department
of  Computer Engineering, Instituto Tecn{\'o}gico de Buenos Aires (ITBA), Ciudad de Buenos Aires,  Argentina e-mail: rramele@itba.edu.ar}% <-this % stops a space
\thanks{Manuscript received ----; revised ---.}}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
250 words objective, methods, results, conclusion and significance
The analysis of Electroencephalographic (EEG) signals is of ulterior importance to aid in the diagnosis of mental disease and to increase our understanding of the brain.  Traditionally, clinical EEG has been analyzed in terms of temporal waveforms, looking at rhythms in spontaneous activity, subjectively identifying troughs and peaks in Event-Related Potentials (ERP), or by studying graphoelements in pathological sleep stages.
Additionally, the discipline of Brain Computer Interfaces requires new methods to decode patterns from non-invasive EEG signals.  This field is developing alternative communication pathways to transmit volitional information from the Central Nervous System. The technology could potentially enhance the quality of life of patients affected by neurodegenerative disorders and other mental illness. This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from images of signal plots.  These features are constructed based on the calculation of histograms of oriented gradients from pixels around the signal plot. It aims to provide a new objective framework to analyze, characterize and classify EEG signal waveforms. The feasibility of the method is outlined by detecting the P300, an ERP elicited by the oddball paradigm of rare events, and implementing an offline P300-based BCI Speller.  The validity of the proposal is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients and an own dataset of healthy subjects.
\end{abstract}

@@

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related to the image scale. In addition, $ x^l(n,c) $ is the point-to-point averaged multichannel EEG signal for the sample point $n$ and for channel $c$. Lastly, $$\bar{x}^l(c) =\frac{1}{n_{max}}\sum_{n=1}^{n_{max}}x^l(n,c)$$ and $$ \hat{\sigma}^l(c) = \Bigg\{ \frac{1}{n_{max}-1}\sum_{n=1}^{n_{max}} \left[  x^l(n,c)-\bar{x}^l(c) \right] ^2 \Bigg\}  ^{\frac{1}{2}}$$ are the mean and estimated standard deviation of $x^l(n,c), 1 \leq n \leq n_{max}$, for each channel $c$.

@@

\begin{equation}
I^{(l,c)}(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \   n \;  \text{and} \;  z_2 = \tilde{x}^l(n,c) + z^l(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

@@

\begin{equation}
z^l(c) = \left \lfloor{ \min_{n} \tilde{x}^l(n,c)  }\right \rfloor
\label{eq:zerolevel}
\end{equation}

@@

In Table~\ref{tab:resultsown} the results obtained for 8 healthy subjects are shown.  It can be observed that the performance is above chance level. It was verified that HIST method has an improved performance at letter identification than SVM that process the signals on a channel by channel strategy (Wilcoxon signed-rank test, $p\_value =  0.004$ for both datasets).

@@

Tables~\ref{tab:resultsalsswlda} and~\ref{tab:resultsownswlda} are presented in order to compare the performance of the HIST method versus a multichannel version of the Stepwise Linear Discriminant Analysis (SWLDA) and SVM classification algorithms for both datasets.  The feature was formed by concatenating all the channels~\cite{Krusienski2006}.  SWLDA is the methodology proposed by the ALS dataset's publisher. Since authors \cite{Riccio2013} did not report the Character Recognition Rate obtained for this dataset, we replicate their procedure and include the performance obtained with the SWLDA algorithm at letter identification.  It was verified for the dataset of ALS patients that it has similar performance  against other methods like SWLDA or SVM, which use a multichannel feature (Quade test with $p\_value=0.55$) whereas for the dataset of healthy subjects significant differences were found (Quade test with $p\_value=0.02$) where only the HIST method achieved a different performance than SVM (multiple comparisons significant difference of level $0.05$).

@@

\begin{figure*}[!t]
\centering
\subfloat[Case I]{\includegraphics[width=15cm]{subject.png}\label{subject8}}
\caption{Ten sample P300 template patches for subjects 8 (A) and 3 (B) of the ALS Dataset.  Downward deflection is positive polarity. The P300 signature waveform is more clearly and consistent characterized for subject 8, whereas for subject 3 the characteristic patterns is more difficult to spot. }
\label{fig:p300templates}
\end{figure*}

Additionally, the stability of the P300 component waveform has been extensively studied in patients with ALS \cite{SellersandEmanuelDonchin2006,TomohiroMadarame2008,Nijboer2009,Mak2012,McCane2015} where it was found that these patients have a stable P300 component, which were also sustained across different sessions.  In line with these results we do not find evidence of a difference in terms of the performance obtained by analyzing the waveforms (HIST) for the group of patients with ALS and the healthy group of volunteers (Mann-Whitney U Test, $p\_value=0.46$). Particularly, the best performance is obtained for a subject from the ALS dataset for which, based on visual observation, the shape of they P300 component is consistently identified.

@@

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)\begin{IEEEbiographynophoto}
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{ramele}}]{Rodrigo Ramele}
currently works at the Centro de Inteligencia Computacional (CIC), ITBA University, Buenos Aires, Argentina. Rodrigo does research in Brain Computer Interfaces, Artificial Intelligence, Signal Processing and Robotics.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{villar}}]{Ana Julia Villar}
was graduated in applied mathematics and got her Ph.D.degree in Robust Statistic at the  Universidad de Buenos Aires, Argentina. She is currently a professor at the Instituto Tecn{\'o}gico de Buenos Aires. Dr. Villar has authored or co-authored several papers in the fields of Brain Computer Interface (BCI), Reinforcement Learning and Robust Statistic.  Her research interests include statistical machine learning, robust statistics and their applications to BCI.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{santos}}]{Juan Miguel Santos}
is the Director of the Centro de Inteligencia Computacional (CiC) of ITBA University in Buenos Aires, Argentina, where he works in three different areas: robotics development, reinforcement learning and brain-computer interfaces.  In the first two lines he works since 1993 developing and constructing robots, national and international robotics conferences and international peer-review journals.  Currently developing a new development line based on SLAM to autonomous robots.
\end{IEEEbiography}

@@
